---
title: "Introduction to drRML package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to drRML package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  dpi = 150,
  fig.path = "vignettes/drRML-"
)
```

## Introduction

`drRML` is the R package to estimate differences in restricted mean lifetimes (RML) of two treatments using pseudo-observations.
Define $T$ and $C$ as the event time and right-censoring times, respectively.
Then we observe $\{Y_i = T_i\wedge C_i,\Delta_i = I(T_i\le C_i),Z_i\}_{i=1}^n$, 
where $Z_i$ is $p-$dimensional baseline covariate vector.
We adopt the pseudo-observation (available in `pseudo` R package)
to replace unknown event times,
$$
\hat \theta^*_{i}
	= n\cdot\int_{0}^{L}\hat{S}(t) dt - (n-1)\cdot\int_{0}^{L}\hat{S}^{-i}(t)dt,
$$
where $L$ is fixed time point for RML and $\hat S(t)$ is the empirical survival estimates, such as the Kaplan-Meier estimator.
Then, the double robust estimator for average causal effect is defined as
$$
\hat\delta_{DR} = \hat\mu_1^{DR}-\hat\mu_0^{DR},
$$
where $\hat\mu_a^{DR}~(a=0,1)$ can be expressed as
$$
\hat\mu_a^{DR} = n^{-1}\sum_{i=1}^n
	\frac{I(A_i=a)}{\tilde \pi_a(Z_i;\hat\gamma)}\hat\theta^*_i
	+ n^{-1}\sum_{i=1}^n \left\{1- \frac{I(A_i=a)}{\tilde\pi_a(Z_i;\hat\gamma)} \right\} m(a,Z_i;\hat\beta).
$$
Here, $\tilde \pi_a(Z_i;\gamma) = a\pi(Z;\gamma) + (1-a) \{1-\pi(Z;\gamma) \}$ 
is the propensity score, estimated by logistic regression or machine learning approaches (e.g., `SuperLearner`, `gbm` packages in R).
In addition, $m(a,Z_i;\beta)$ is the outcome regression model, estimated by
linear regression or machine learning approaches (e.g., `SuperLearner` in R).
See Choi et al. (2022+) for more detailed description of the method.



## Installation
```{r}
library(devtools)
devtools::install_github(repo='taehwa015/drRML')
```


## Usage

The `drRML` package provides a double-robust estimator for differences in restricted mean lifetimes using pseudo-observations.

Below example is one of simulation set-up in Choi et al. (2022+).
Simulated event time variable involves three covariates
$Z=(Z_1,Z_2,Z_3)'$ distributed as multivariate normal with mean zero, unit variance, corr$(Z_1,Z_3)=0.2$, with all other pairwise correlations set to be 0. The treatment variable $A\in\{0,1\}$ is created from a logistic regression model with $P(A=1|Z)=\text{expit}(-0.5 Z_1-0.5 Z_2)$, where $\text{expit}(x)=1/(1+e^{-x})$.
Then the event time $T$ is generated from an exponential distribution with rate $\exp(-3-Z_1-0.9 Z_2-Z_3)$ for treatment $A=1$, and $\exp(-2.5-1.5Z_1-Z_2-0.7Z_3)$ for $A=0$.
The censoring time $C$ follows an exponential with rate $e^{-4.5}$, yielding approximately 25\% censoring.
```{r}
simdata = function(n, tau) {
  require(mvtnorm)
  expit = function(x) exp(x) / (1 + exp(x))
  Sigma = diag(rep(1, 3))
  Sigma[1, 3] = Sigma[3, 1] = 0.2
  Z = rmvnorm(n,
              mean = rep(0, 3),
              sigma = Sigma,
              method = "chol")
  Z1 = Z[, 1]
  Z2 = Z[, 2]
  Z3 = Z[, 3]
  # trt indicator (Bernoulli)
  A = rbinom(n, 1, expit(-0.5 * Z1 - 0.5 * Z2))# Z1 and Z2 are involved in trt assignment
  par.t = exp((-2.5 - 1.5 * Z1 - 1 * Z2 - 0.7 * Z3) * (A == 0) +
                (-3 - 1 * Z1 - 0.9 * Z2 - 1 * Z3) * (A == 1))
  T = rexp(n, par.t)                   # true surv time
  C = rexp(n, tau)                     # independent censoring
  # tau as a controller of censoring rate
  Y = pmin(T, C)                       # observed time
  delta = ifelse(T <= C, 1, 0)         # censoring indicator
  simdata = data.frame(Y = round(Y, 5), delta, A, Z1, Z2, Z3, T)
  simdata[order(Y), ]
}
L = 10       
n = 600      
tau = exp(-4.5)
truepar = ifelse(L == 10, 0.871, 1.682)

set.seed(123)
dt = simdata(n, tau)
Y = dt$Y
Censor = dt$delta
A = dt$A
X = dt[, 4:6]
```

Once we estimate $\pi(Z;\gamma)$ and $m(a,Z;\beta)$
from logistic and gaussian regressions, respectively,
`dr_rml` can be used by specifying `PS = "logit"` and
`Reg = "lm"`,
while 10 bootstrapped samples are considered for standard error estimates of ACE (`nboot = 10`).
```{r}
dr_rml(Y = Y, Censor = Censor, A = A, Xps = X, Xreg = X, 
       L = L, PS = "logit", Reg = "lm", nboot = 10)$ace
```
Even if we assume the misspecified propensity score by omiting $Z_2$,
proposed estimator still consistent.
```{r}
dr_rml(Y = Y, Censor = Censor, A = A, Xps = X[,-2], Xreg = X, 
       L = L, PS = "logit", Reg = "lm", nboot = 10)$ace
```

For flexible estimation, we may alternatively use other machine learning methods. For example, we estimate the propensity score model from super learner
(`PS = "SL"`)
```{r}
dr_rml(Y = Y, Censor = Censor, A = A, Xps = X, Xreg = X,
       L = L, PS = "SL", Reg = "lm", nboot = 10)$ace
```

Below example is the GSE6532 data application in the article.
```{r}
library(drRML)
data(gse)
Y = gse$Y
Censor = gse$Censor
A = gse$trt
X = gse[, 3:6]
L = 365 * 5
dr_rml(Y = Y, Censor = Censor, A = A, Xps = X, Xreg = X,
       L = L, PS = "logit", Reg = "lm", nboot = 10)
```



